Supplementary Information: Vulnerability of tree growth

### 2025-03-21 doc/manuscript_hkk_drought_sensitivity.Rmd
moved table 1 after hypothesis paragraph
Update manuscript_hkk_drought_sensitivity.Rmd
- added/ organized some intro content (including copying some content from grant proposal)

<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- Climate anomalies like drought can lead to stand-level growth reductions, along with increased mortality [@bennett_sensitivity_2023; @bauman_tropical_2022].  
<!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
<!-- - Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017]. -->
- Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.  
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns.
- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).   
- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->

>>>>

<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
- Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].<!--copied from our grant proposal - something like this is more specific and perhaps more compelling-->
- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- However, tree growth is increasingly vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022]. <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
- (We don't know very much about woody growth responses, as few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].)
- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.  
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns.
- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
- Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).   
- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->
